Think deep for reef resilience

### 2025-09-04 paper/paper_draft.qmd
Update paper citations in discussion and format doc

| Global Climate Model source | ECS value (°C) | Interpreted relative severity | Results found in |
|------------------|------------------|-------------------|------------------|
| ACCESS-CM2 | 4.66 [@grose2023] | High | [@grier2025] |
| EC-Earth3-Veg | 4.30 [@wyser2020] | High | [@grier2025] |
| GFDL-CM4 | 4.1 [@sentman2025] | High | [@grier2025] |
| ACCESS-ESM1-5 | 3.87 [@ziehn2020] | Intermediate | Main text |
| NorESM2-MM | 2.5 [@seland2020] | Low | [@grier2025] |

@@

While 99% accuracy suggests the classifier is overfit to the training data, the 66% test accuracy indicates a meaningful relationship between modelled reef characteristics and their assigned clusters has been captured. A 3-fold cross-validation was also conducted, arriving at a similar 67% accuracy rate. The consistent level of accuracy across test and cross-validation demonstrates the generalised performance of the RF classifier is reliable. Furthermore, F1 performance scores for Low, Medium, and High clusters were determined to be 0.7, 0.5 and 0.65 respectively, indicating balanced performance across the cluster categories. Lower performance with the Medium cluster reflects the inherent difficulty of classifying transitional states in natural systems. For management applications, this level of performance provides valuable decision support when combined with expert knowledge and the model's probabilistic outputs.

@@

These findings support the hypothesis that reefs at greater depths will exhibit characteristics of refugia and more prolonged structural resilience under climate change compared to shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024] suggest that rubble instability is inversely correlated with reef depth, possibly limiting juvenile recruitment capacity of shallow reefs. A caveat to these depth-resilience findings is the quality of the bathymetry data used to calculate median reef depth. As detailed previously, estimates of median reef depth are derived from satellite imagery (at $10m^2$ resolution) which can be subject to high degrees of uncertainty, both in terms of the estimated depth at for each pixel as well as their alignment to the established GBRMPA reef outlines. Additionally, we do not incorporate the light-based

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this investigation. We model coral colonies as circular diameter values that are tracked over timesteps through growth. Additionally, we do not model the direct influence of light intensity on coral growth or the negative influence of depth on coral growth rates due to reduced available light levels. [@baker1975; @huston1985; @ruiz-diaz2022]. If included in this investigation it is possible that reef recovery ability, via regrowth, would be reduced at greater depths, reducing the positive effect of depth refugia for temporal resilience. Additionally, reduced growth rates at increased depths may reduce the carbonate budget maintenance of reefs due to lower rates of CaCO~3~ accretion. Due to the availability of quality future projected data we do not consider other important impacts such as waves, cyclones, currents and Crown of Thorns Starfish (CoTS) on coral colony survival and bleaching [@massel1993; @deaker2022].

@@

| Metric | Used in | Definition | Justification |
|------------------|------------------|------------------|------------------|
| Percentage coral cover time series | Timeseries clustering, Carbonate budget analysis | Total coral cover as a percentage of reef area | To summarise coral population health over time |
| Median depth | Time series cluster assessment Carbonate budget analysis | Reef median depth as calculated with zonal raster analysis | Depth is a critical form of environmental refugia from high thermal stress |
| Mean time series DHW | Time series cluster assessment | Reef mean DHW value across time series | Resilient reefs may experience reduced surface DHWs |
| Source - to - sink connectivity ratio | Time series cluster assessment | Ratio of reef outgoing to incoming connectivity | Reefs with greater source than sink connectivity may have lower recovery |
| Weighted incoming connectivity | Time series cluster assessment, Carbonate budget analysis | Incoming connectivity strength weighted by mean time series coral cover of source reefs | High incoming larval flow assists reef recovery |
| Weighted outgoing connectivity | Time series cluster assessment | Outgoing connectivity strength weighted by mean coral cover | High outgoing larval flow may lower recovery capacity of the target reef |
| Carrying capacity | Time series cluster assessment | Reef habitable area \[m^2^\] | Captures the size of the viable reef extent |

>>>>

| Global Climate Model source | ECS value (°C)     | Interpreted relative severity | Results found in |
|-----------------------|-----------------------|------------------------|-----------------------|
| ACCESS-CM2                  | 4.66 [@grose2023]  | High                          | [@grier2025]     |
| EC-Earth3-Veg               | 4.30 [@wyser2020]  | High                          | [@grier2025]     |
| GFDL-CM4                    | 4.1 [@sentman2025] | High                          | [@grier2025]     |
| ACCESS-ESM1-5               | 3.87 [@ziehn2020]  | Intermediate                  | Main text        |
| NorESM2-MM                  | 2.5 [@seland2020]  | Low                           | [@grier2025]     |

@@

While 99% accuracy suggests the classifier is overfit to the training data, the 66% test
accuracy indicates a meaningful relationship between modelled reef characteristics and their
assigned clusters has been captured. A 3-fold cross-validation was also conducted, arriving
at a similar 67% accuracy rate. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier is reliable.
Furthermore, F1 performance scores for Low, Medium, and High clusters were determined to be
0.7, 0.5 and 0.65 respectively, indicating balanced performance across the cluster
categories. Lower performance with the Medium cluster reflects the inherent difficulty of
classifying transitional states in natural systems.

@@

These findings support the hypothesis that reefs at greater depths will exhibit
characteristics of refugia and more prolonged structural resilience under climate change
compared to shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024] suggest
that rubble instability is inversely correlated with reef depth, possibly limiting juvenile
recruitment capacity of shallow reefs. A caveat to these depth-resilience findings is the
quality of the bathymetry data used to calculate median reef depth. As detailed previously,
estimates of median reef depth are derived from satellite imagery (at $10m^2$ resolution)
which can be subject to high degrees of uncertainty, both in terms of the estimated depth at
for each pixel as well as their alignment to the established GBRMPA reef outlines.
Additionally, we do not incorporate the light-based decreases in coral growth rates as depth
increases [@baker1975; @huston1985; @ruiz-diaz2022].

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as circular diameter values that are tracked over
timesteps through growth. Additionally, we do not model the direct influence of light
intensity on coral growth or the negative influence of depth on coral growth rates due to
reduced available light levels [@baker1975; @huston1985; @ruiz-diaz2022]. If included in
this investigation it is possible that reef recovery ability, via regrowth, would be reduced
at greater depths, reducing the positive effect of depth refugia for temporal resilience.
Additionally, reduced growth rates at increased depths may reduce the carbonate budget
maintenance of reefs due to lower rates of CaCO~3~ accretion. Due to the availability of
quality future projected data we do not consider other important impacts such as waves,
cyclones, currents and Crown of Thorns Starfish (CoTS) on coral colony survival and
bleaching [@massel1993; @deaker2022].

@@

| Metric | Used in | Definition | Justification |
|-----------------------|-----------------------|-----------------------|-----------------------|
| Percentage coral cover time series | Timeseries clustering, Carbonate budget analysis | Total coral cover as a percentage of reef area | To summarise coral population health over time |
| Median depth | Time series cluster assessment Carbonate budget analysis | Reef median depth as calculated with zonal raster analysis | Depth is a critical form of environmental refugia from high thermal stress |
| Mean time series DHW | Time series cluster assessment | Reef mean DHW value across time series | Resilient reefs may experience reduced surface DHWs |
| Source - to - sink connectivity ratio | Time series cluster assessment | Ratio of reef outgoing to incoming connectivity | Reefs with greater source than sink connectivity may have lower recovery |
| Weighted incoming connectivity | Time series cluster assessment, Carbonate budget analysis | Incoming connectivity strength weighted by mean time series coral cover of source reefs | High incoming larval flow assists reef recovery |
| Weighted outgoing connectivity | Time series cluster assessment | Outgoing connectivity strength weighted by mean coral cover | High outgoing larval flow may lower recovery capacity of the target reef |
| Carrying capacity | Time series cluster assessment | Reef habitable area \[m^2^\] | Captures the size of the viable reef extent |
